---
title: "Data Wrangling Exercise 1: Basic Data Manipulation"
subtitle: "Springboard: Foundations of Data Science"
author: "Antoine Beauchamp"
date: "January 13th, 2017"
output: html_document
---

##Introduction

Let's begin by clearing the working environment and load the require libraries. 

``` {r}
rm(list=ls())

suppressMessages(library(dplyr))
library(tidyr)
library(readr)
```

Next we will make sure that we are in the correct working directory. 

``` {r}
path_to_wd <- file.path("~","Documents","Work","DataScience","Springboard","FoundationsofDataScience","Section3_DataWrangling","DataWrangling_Exercise1")
setwd(path_to_wd)
rm(path_to_wd)
```

##Section 0: Import Data 

Here we read in the .csv data into a tibble data frame, `refine_data`. Let's also assign variable names. 
``` {r}
colnames_vec=c("company","product_code_number","address","city","country","name")
refine_data <- suppressMessages(tbl_df(read_csv("refine_original.csv", col_names=c("company","product_code_number","address","city","country","name"))))
#Drop unnecessary name row
refine_data <- refine_data[-1,]
```

A few functions to take a look at the data:
``` {r}
class(refine_data)
glimpse(refine_data)
#With the tibble format we can just print the data in a clean manner:
refine_data
```


##Section 1: Clean Up Brand Names 

Let's correct the name data for each of the companies. 
``` {r}
#Correct the data for each of the companies. (Is there a better way of doing this? I can't imagine that this would be helpful for a large dataset. )
d1 <- refine_data %>% mutate(company= gsub("^([Pp][Hh]|[Ff]).+[Pp][Ss]$", "Phillips", refine_data$company)) 
d2 <- refine_data %>% mutate(company=gsub("^[Aa][Kk].*[Zz]([Oo]|0)$", "Akzo", d1$company))
d3 <- refine_data %>% mutate(company=gsub("^[Vv].+[Hh].+en$", "Van Houten", d2$company))
refine_data1 <- refine_data %>% mutate(company=gsub("^[Uu].*er", "Unilever", d3$company))


#Check the result to be sure
refine_data1$company
```  
Looks good!


##Section 2: Separate Product and Code Number

Let's separate the product and code number with a simple `separate()` statement:

``` {r}
refine_data2 <- refine_data1 %>% separate(product_code_number,into=c("product_code","product_number"),sep="-")

#Check to be sure
refine_data2
```

##Section 3: Add Product Category

Next we want to add a product category to correspond to the product code. First let's define a function that take a vector of product numbers and returns a vector of product categories:

``` {r}
prodcategory_func <- function(invec){
  newvec = rep(0, length(invec))
  for (i in 1:length(invec)){
    if (invec[i]=='p'){
      newvec[i]='Smartphone'
    } else if (invec[i]=='v') {
      newvec[i]='TV'
    } else if (invec[i]=="x") {
      newvec[i]="Laptop"
    } else if (invec[i]=="q") {
      newvec[i]="Tablet"
    }
  }
    return(newvec)
}
```
Now that we have this, we will use `mutate()` to generate the product category column.

``` {r}
refine_data3 <- refine_data2 %>% mutate(product_category=prodcategory_func(refine_data2$product_code))

#Check to be sure
refine_data3 %>% as.data.frame() %>% head()
```


## Section 4: Add Full Address 
Let's combine the `address`, `city`, and `country` variables using a simple `unite()` statement:     
``` {r}     
refine_data4 <- refine_data3 %>% unite(full_address, address, city, country, sep=",")
 
refine_data4 %>%  as.data.frame() %>% head()
```

## Section 5: Create dummy variables for company and product

Now we want to create dummy binary variables for each of the companies and product types. We can do this with consecutive `mutate()` statements:

``` {r}
#First the companies
refine_data5 <- refine_data4 %>% 
  mutate(company_phillips=sapply(refine_data4$company, function (x) {if (x=="Phillips") {1} else {0}}))  %>% 
  mutate(company_akzo=sapply(refine_data4$company, function (x) {if (x=="Akzo") {1} else {0}})) %>% 
  mutate(company_van_houten=sapply(refine_data4$company, function (x) {if (x=="Van Houten") {1} else {0}})) %>% 
  mutate(company_unilever=sapply(refine_data4$company, function (x) {if (x=="Unilever") {1} else {0}}))

#Second the products
refine_clean <- refine_data5 %>% 
  mutate(product_smartphone=sapply(refine_data4$product_category, function (x) {if (x=="Smartphone") {1} else {0}})) %>%
  mutate(product_tv=sapply(refine_data4$product_category, function (x) {if (x=="TV") {1} else {0}})) %>%  
  mutate(product_laptop=sapply(refine_data4$product_category, function (x) {if (x=="Laptop") {1} else {0}})) %>%
  mutate(product_tablet=sapply(refine_data4$product_category, function (x) {if (x=="Tablet") {1} else {0}}))
  

refine_clean %>% as.data.frame() %>% head()

```

  
  
  
  
##Section 6: Write to File 
  
Finally let's finish this off by writing the new clean data to a .csv file.
``` {r}  
write_csv(refine_clean, "refine_clean.csv")
```
  

